Blaming the Typeless: \\ Scalable, Human-Centric Python Localization

### 2018-08-22 paper/main.tex
paper updates

We propose an algorithm that takes an input a Python program fragment that 
produces an uncaught runtime exception such as a type error. Via machine
learning over a carefully-selected set of program-relevant features, we
produce the expression most likely implicated in the exception according to
our learned model. We evaluate on an entire year of buggy Python 3 program
executions on PythonTutor.com --- over 270,000 instances --- measuring
accuracy with respect to the human actions as the ground truth and
comparing against the vanilla Python interpreter as a baseline.

@@

One reasonable fix (and the one that the programmer actually used in this
case) is to add \lt{int()} cast on line 2. However, another possible
``fix'' would be to \textit{remove} the \lt{int()} cast from line 1, in
which case both variables hold string values at runtime, and the addition
on line 4 would be interpreted as string concatenation. While the second
fix creates a well-typed program, it is less likely to correspond to
developer intent, and is thus less likely to be helpful as a debugging aid.

@@

Our algorithm first extracts static and dynamic features from Python
program (Section~\ref{sec-features}). Next, using a labeled training
corpus, we learn a machine learning model over those features
(Section~\ref{sec-model}). Once the model has been learned, we localize
faults in new Python programs by extracting their features and applying the
model.

Drawing inspiration from localization algorithms such as
Nate~\cite{learning-to-blame} and the natural language processing term
frequency vector (or ``bag of words'') model~\cite{FIXME}, we represent
each buggy program as a ``bag of abstracted terms''. A \emph{term} is
either a statement of expression.

@@

For each term in the program we compute static, dynamic, and contextual
features. While some features can be computed directly from syntactic
information, others are only only accessible dynamically. In a
dynamically-typed language such as Python, this includes types of each
expression. Since programs are highly structured and expressions gain
meaning in relation to their surroundings, we also include contextual
features.

While static features can be computed via parsing, dynamic features require
executing the program fragment. To calculate dynamic features, we first
convert each buggy program to an equivalent program that is closer to ANF.
\fixme{WRW has no idea what ``ANF'' means in this context, and this is the
first and only time it appears in the paper text. Someone else should
rewrite this.} We define ANF to be FIXME and ``closer'' to mean FIXME in
this context. We then run the resulting program through the PythonTutor
backend~\cite{FIXME}.  This backend is based on the Python debigging
library BDB~\cite{FIXME}; it returns a trace of the program along with the
state of the heap at each execution step. That trace is used to compute
dynamic features.

@@

To overcome this issues in the ``early return'' or ``early break'' case
(\figref{early-break}), we check if a break, return, or other statement for
escaping structured control flow is present inside of a conditional
statement. We then add dependencies in the execution trace's dependency graph
between the enclosing conditional and the statements that would have been skipped
by the break or return. While this heuristic is effective in practice, it
does not overcome all related problems: we thus treat dynamic slice
information as one of many features rather than as a hard constraint.
This boolean feature tracks whether or not the term is part of the final
dynamic slice.

@@

\subsubsection{Exception Type (Dynamic)} The type of error thrown by the program
contains useful information its localization. In particular, a term may be
more or less suspicious depending on the ultimate uncaught exception. As an
example, a \lt{v / 0} term may be more suspicious if the uncaught exception
being considered is \lt{DivisionByZero}. We encode the exception type as
as a categorical feature. 
% to \emph{every} vector derived from that program.

\fixme{The original draft contained this text: ``For example, if asked how
likely a random `0` is of being a bug, you'd give a higher answer if you
knew that the program eventually crashes with a division by zero exception.''
This sort of rhetorical communication with the reader is \emph{too
informal for scientific writing}. I am not saying that I disagree with the
underlying argument. I am saying that you cannot really use the Socratic
method or the ``Appeal to Common Practice'' logical fallacy or the ``No
True Scotsman'' logical fallacy in scientific writing. If you wrote the
quoted text, read this FIXME carefully, ask me if you have any questions
about sloppy vs. formal arguments, and then delete this FIXME. --- Wes}

@@

As an example, consider the \lt{0} terms in both \lt{x = 0} and \lt{ x/ 0}.
Both instances of \lt{0} have the same syntactic form, size and type
(etc.). We may prefer to implicate the \lt{0} in \lt{x / 0} as suspicious,
especially for beginner-written programs, but cannot distinguish it without
surrounding contextual information. Data structures such as abstract syntax
trees and control flow graphics capture such contextual information, but
are not immediately applicable to machine learning.

We desire to encode such information while retaining the use of scalable,
accurate off-the-shelf machine learning algorithms that operate on feature
vectors. We thus propose to embed contextual information in a vector,
borrowing insights from standard approaches in machine learning. We
associated with term additional features that correspond to the features of
its parent and child nodes. For representational regularity, we always
model three children; terms without parents or children are given special
values (e.g., zero or NotApplicable) for those contextual features.

@@

\fixme{
\emph{Models} The following descriptions of Logistic Regression, Decision Trees,
and MLP will have to be redone as they are copied from Learning to Blame.
}

@@

@@

Our raw data consist of every Python 3 program that a user executed on
PythonTutor.com~\cite{Guo2013-vu} (not in ``live'' mode) during 2017, other
than those with syntax errors or undefined variables.  Each program which
crashes (throws an uncaught Python exception) is paired with the next
program (by the same user) that does not crash, under the assumption that
the latter is the fixed version of the former. We discard pairs where the
difference between crashing and fixed versions is too high (more than a
stddev above average), since these are most likely to be violations of that
assumption (i.e., the program that does not crash is unrelated to the
crashing program). We also discard submissions that are out of scope given
PythonTutor's policies (e.g., such as very long-running executions or the
use of forbidden libraries).

@@

We use the terms changed involved in the fixes by actual users as our
ground truth and training information.  Many PythonTutor users use the site
in an iterative fashion: they start out by writing a program that crashes,
and then edit it until it no longer crashes. Our dataset contains only only
those crashing programs for which the same user later submitted a program
that did not crash. We compute a tree-diff \cite{tree-diff} between the
originaly, buggy submission and the first fixed submission. We define 
the ground truth correct answer to be the set of terms in the crashing
program that also appear in the diff. We discuss the implications of this
choice in \secref{threats}.

@@

For these evaluations, we employ machine learning algorithm FIXME, which
provided the best balance of FIXME and FIXME (time? accuracy?) on our
training set.

@@

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
    ybar,
    symbolic x coords={Baseline,Top-1,Top-3},
    xtick=data,
    ymin=0,
    ylabel=Model Accuracy,
    enlarge x limits=0.5,
    legend style={at={(0.5,-0.15)},anchor=north}
]
    \addplot table[x=scoreName, y=score, col sep=comma]{fault-localization.csv};
\end{axis}
\end{tikzpicture}
\caption{Baseline is the normal Python interpreter. Other bars are from decision
trees on the largest dataset (272534 pairs). TODO: also train for Top-2}
\label{fig:full-dataset-acc}
\end{figure}

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
    ybar,
    symbolic x coords={noContext, noTypes, all},
    xtick=data,
    ymin=0,
    ylabel=Model Accuracy,
    enlarge x limits=0.5,
    legend style={at={(0.5,-0.15)},anchor=north}
]
    \addplot table[x=scoreName, y=score, col sep=comma]{removing-features-2.csv};
\end{axis}
\end{tikzpicture}
\caption{Accuracy when features are removed, based on decision trees on a random subset of 20000 pairs.
TODO: Top-2}
\label{fig:removing-features}
\end{figure}

prereq: define accuracy

Goal: Establish that our algorithm works. This is the most basic question,
but also the most relevant. If the reader takes away nothing else, they
should conclude that our algorithm is effective.

PLOT: See Figs. \ref{fig:full-dataset-acc} and \ref{fig:removing-features}

@@

\begin{table}[]
\begin{tabular}{lll}
Name & Category & Gini Importance \\ \bottomrule
Size-P & Contextual (Syntactic)               & 0.112 \\
Stmt-Constr-C3= & Contextual (Syntactic)      & 0.061 \\
Expr-Constr-P=List & Contextual (Syntactic)   & 0.055 \\
PythonBlame & Dynamic (Error location)        & 0.039 \\
Type=unknown & Dynamic (Type)                 & 0.037 \\
Type-P=unknown & Contextual (Type)            & 0.037 \\
PythonMsg=IndexError & Dynamic (Error message)& 0.033 \\
Size & Syntactic                              & 0.028 \\
Expr-Constr-P=Dictionary & Contextual (Syntactic) & 0.027 \\
Size-C1 & Contextual (Syntactic)              & 0.024 \\
Expr-Constr=Var & Syntactic                   & 0.020 \\
Type-C1=unknown & Contextual (Type)           & 0.020 \\
\toprule
\end{tabular}
\caption{Feature predictive power (for a Top-3 Decision Tree
learned on the entire dataset).
TODO: add
better-known measure like ANOVA or ReliefF. TODO: base on Top-1 instead?}
\label{tab-feature-predictive-power}
\end{table}

@@

prereq: define predictive power (leave-one-out, leave-one-in, Relief-F,
ANOVA, whatever)

PLOT: See Table \ref{tab:feature-predictive-power}

Use as a rough guide: Table 3 on Page 8 of
https://web.eecs.umich.edu/~weimerw/p/weimer-icsm2010.pdf

Goal: Establish that we are smart for including all of these features. Note
which individual features or feature categories were (not) included in
previous work.

Give a simple narrative about our effectiveness: implicitly, we were smart
for deciding to include these features, better ingredients make better
results, we have great results.

TODO: what about turning on/off slicing?

TODO: what about turning on/off types? (is that separate from slicing?)

@@

%In Nate there were 20 different functions. All the programs were doing the
%same 20 functions.
%
%Here, people are doing anything they want. Random stuff going on over here.
%
%We want to show that we get high accuracy *over a heterogenous set of
%programs*. Leon?
%
%prereq: cluster the programs based on similarity. Using agglomerative
%clustering plus python edit distance. How good the performance will be?
%right now it creates a clustering tree. binary tree, based on min distance.
%can give it a threshhold for distance (elements w/in a group). don't forget
%to rename all variables to X, perhaps normalize (edit distance size).
%
%leon: could also cluster by inconsistency coefficient
%
%PLOT Type: line graph (standard plot)
%
%PLOT X AXIS: clustering distance parameter
%
%PLOT Y AXIS: number of different clusters found
%
%PLOT LINES: line 1 is ``Nate programs'', line 2 is ``25\%, at random, of our dataset''
%
%Goal: show that ``no matter how you decide program differences'', our
%dataset is always more diverse/heterogeneous than is previous work. So our
%success is amazing.

@@

As a measure of the diversity of our dataset, we clustered the programs in
our dataset and the NATE dataset, then compared the number of resultant clusters.
To cluster programs, we transform the program ASTs into flat strings, and then
perform agglomerative clustering on them, with string edit distance as a distance
metric.

>>>>

We propose an algorithm that takes as input a Python program fragment that
produces an uncaught runtime exception such as a type error. Via machine
learning over a carefully-selected set of program-relevant features, we
produce the expression most likely implicated in the exception according to
our learned model. We evaluate on an entire year of buggy Python 3 program
executions on PythonTutor.com --- over 270,000 instances --- measuring
accuracy with respect to the human actions as the ground truth and
comparing against the vanilla Python interpreter as a baseline.

@@

One reasonable fix (and the one that the programmer actually used in this
case) is to add an \lt{int()} cast on line 2. However, another possible
``fix'' would be to \textit{remove} the \lt{int()} cast from line 1, in
which case both variables hold string values at runtime, and the addition
on line 4 would be interpreted as string concatenation. While the second
fix creates a well-typed program, it is less likely to correspond to
developer intent, and is thus less likely to be helpful as a debugging aid.

@@

Our algorithm first extracts static and dynamic features from a Python
program (Section~\ref{sec-features}). Next, using a labeled training
corpus, we learn a machine learning model over those features
(Section~\ref{sec-model}). Once the model has been learned, we localize
faults in new Python programs by extracting their features and applying the
model.

Drawing inspiration from localization algorithms such as
Nate~\cite{learning-to-blame} and the natural language processing term
frequency vector (or ``bag of words'') model~\cite{FIXME}, we represent
each buggy program as a ``bag of abstracted terms''. A \emph{term} is
either a statement or expression.

@@

For each term in the program we compute static, dynamic, and contextual
features. While some features can be computed directly from syntactic
information, others are only accessible dynamically. In a
dynamically-typed language such as Python, this includes types of each
expression. Since programs are highly structured and expressions gain
meaning in relation to their surroundings, we also include contextual
features.

While static features can be computed via parsing, dynamic features require
executing the program fragment. To calculate dynamic features, we first
convert each buggy program to an equivalent program in a form similar to
A-normal form \cite{anf}: all intermediate expressions are split out and
assigned to new temporary variables.
We then run the resulting program through the PythonTutor
backend~\cite{Guo2013-vu}.  This backend is based on the Python debugging
library BDB~\cite{FIXME}; it returns a trace of the program along with the
state of the heap at each execution step. That trace is used to compute
dynamic features.

@@

To overcome these issues in the ``early return'' or ``early break'' case
(\figref{early-break}), we check if a break, return, or other statement for
escaping structured control flow is present inside of a conditional
statement. We then add dependencies in the execution trace's dependency graph
between the enclosing conditional and the statements that would have been skipped
by the break or return. While this heuristic is effective in practice, it
does not overcome all related problems: we thus treat dynamic slice
information as one of many features rather than as a hard constraint.
This boolean feature tracks whether or not the term is part of the final
dynamic slice.

@@

\subsubsection{Exception Type (Dynamic)} The type of error thrown by the program
contains useful information for its localization. In particular, a term may be
more or less suspicious depending on the ultimate uncaught exception. As an
example, every division term may be more suspicious if the uncaught exception
being considered is \lt{DivisionByZero}. We encode the exception type as
a categorical feature.
% to \emph{every} vector derived from that program.

@@

As an example, consider the \lt{0} terms in both \lt{x = 0} and \lt{x / 0}.
Both instances of \lt{0} have the same syntactic form, size and type
(etc.). We may prefer to implicate the \lt{0} in \lt{x / 0} as suspicious,
especially for beginner-written programs, but cannot distinguish it without
surrounding contextual information. Data structures such as abstract syntax
trees and control flow graphics capture such contextual information, but
are not immediately applicable to machine learning.

We desire to encode such information while retaining the use of scalable,
accurate off-the-shelf machine learning algorithms that operate on feature
vectors. We thus propose to embed contextual information in a vector,
borrowing insights from standard approaches in machine learning. We
associate with each term additional features that correspond to the features of
its parent and child nodes. For representational regularity, we always
model three children; terms without parents or children are given special
values (e.g., zero or NotApplicable) for those contextual features.

@@

\fixme{
\emph{Models} The following descriptions of Logistic Regression, Decision Trees,
Random Forests, and MLP will have to be redone as they are copied from Learning to Blame.
}

@@

\emph{Random Forests}
%
Random forests improve generalization by training an
\emph{ensemble} of distinct decision trees and using a majority
vote to make a prediction.
%
The agreement among the trees forms a natural
confidence score.
%
Since each classifier in the ensemble is a decision tree, this still
allows for complex and expressive classifiers.

The training process involves taking $N$ random subsets of the training
data and training a separate decision tree on each subset --- the
training process for the decision trees is often modified slightly to
reduce correlation between trees, by forcing each tree to pick features
from a random subset of all features at each node.
%
The diversity of the underlying models tends to make random forests less
susceptible to the overfitting, but it
also makes the learned model more difficult to interpret.

@@

Our raw data consist of every Python 3 program that a user executed on
PythonTutor.com~\cite{Guo2013-vu} (not in ``live'' mode) during 2017, other
than those with syntax errors or undefined variables.  Each program which
crashes (throws an uncaught Python exception) is paired with the next
program (by the same user) that does not crash, under the assumption that
the latter is the fixed version of the former. We discard pairs where the
difference between crashing and fixed versions is too high (more than a
stddev above average), since these are most likely to be violations of that
assumption (i.e., the program that does not crash is unrelated to the
crashing program). We also discard submissions that are out of scope given
PythonTutor's policies (e.g., very long-running executions or the
use of forbidden libraries).

@@

We use the terms changed involved in the fixes by actual users as our
ground truth and training information.  Many PythonTutor users use the site
in an iterative fashion: they start out by writing a program that crashes,
and then edit it until it no longer crashes. Our dataset contains only
those crashing programs for which the same user later submitted a program
that did not crash. We compute a tree-diff \cite{tree-diff} between the
originaly, buggy submission and the first fixed submission. We define
the ground truth correct answer to be the set of terms in the crashing
program that also appear in the diff. We discuss the implications of this
choice in \secref{threats}.

@@

For these evaluations, we employ decision trees as our machine learning
algorithm, since they provided the best accuracy on our data given the
available training time.

@@

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
    ybar,
    symbolic x coords={Baseline,Top-1,Top-2,Top-3},
    xtick=data,
    ymin=0,
    ylabel=Model Accuracy,
    enlarge x limits=0.3,
    legend style={at={(0.5,-0.15)},anchor=north}
]
    \addplot table[x=scoreName, y=score, col sep=comma]{fault-localization.csv};
\end{axis}
\end{tikzpicture}
\caption{Baseline is the normal Python interpreter. Other bars are from decision
trees on the largest dataset (272534 pairs).}
\label{fig-full-dataset-acc}
\end{figure}

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
    ybar,
    symbolic x coords={Baseline,Top-1,Top-3},
    xtick=data,
    ymin=0,
    ylabel=Model Accuracy,
    enlarge x limits=0.3,
    legend style={at={(0.5,-0.15)},anchor=north}
]
    \addplot table[x=scoreName, y=score, col sep=comma]{fault-localization-random-forest.csv};
\end{axis}
\end{tikzpicture}
\caption{Baseline is the normal Python interpreter. Other bars are from random
forests on the largest dataset (272534 pairs). TODO: get top-2, then change text
to discuss this figure instead of decision-tree version}
\label{fig-full-dataset-acc-random-forest}
\end{figure}

We first use the full dataset to train and test decsion trees optimized for
Top-1, Top-2, and Top-3 accuracy. As shown in \figref{full-dataset-acc},
these produce a correct answer 55, 68, and
75\% of the time (to two significant figures). As a baseline, we determine that
the expression blamed by the standard Python interpreter (i.e. the one that
causes it to crash) is only changed by the user 30\% of the time. Thus our
most directly comparable model, the one for Top-1, significantly outperforms
this baseline, and as we then exercise our power to give a ranked output rather
than a single answer, our margin gets even better.

% Goal: Establish that our algorithm works. This is the most basic question,
% but also the most relevant. If the reader takes away nothing else, they
% should conclude that our algorithm is effective.

@@

\begin{table}[]
\begin{tabular}{lll}
Name & Category & Gini Importance \\ \bottomrule
Size-P & Contextual (Syntactic)               & 0.112 \\
Stmt-Constr-C3= & Contextual (Syntactic)      & 0.061 \\
Expr-Constr-P=List & Contextual (Syntactic)   & 0.055 \\
PythonBlame & Dynamic (Error location)        & 0.039 \\
Type=unknown & Dynamic (Type)                 & 0.037 \\
Type-P=unknown & Contextual (Type)            & 0.037 \\
PythonMsg=IndexError & Dynamic (Error message)& 0.033 \\
Size & Syntactic                              & 0.028 \\
Expr-Constr-P=Dictionary & Contextual (Syntactic) & 0.027 \\
Size-C1 & Contextual (Syntactic)              & 0.024 \\
Expr-Constr=Var & Syntactic                   & 0.020 \\
Type-C1=unknown & Contextual (Type)           & 0.020 \\
\toprule
\end{tabular}
\caption{Feature predictive power (for a Top-3 Decision Tree
learned on the entire dataset).}
\label{tab-feature-predictive-power}
\end{table}

\begin{table}[]
\begin{tabular}{lll}
Name & Category & F-score \\ \bottomrule
Stmt-Constr-C3= & Contextual (Syntactic) & 69 \\
Size & Syntactic                         & 63 \\
PythonBlame & Dynamic (Error location)   & 50 \\
Type-P=LIST & Contextual (Type)          & 36 \\
Type-C1=bool & Contextual (Type)         & 30 \\
Type-P=unknown & Contextual (Type)       & 28 \\
Type-C1= & Contextual (Type)             & 25 \\
Size-C3 & Contextual (Syntactic)         & 20 \\
Op-Constr-C1= & Contextual (Syntactic)   & 20 \\
Size-C1 & Contextual (Syntactic)         & 18 \\
Type-P=DICT & Contextual (Type)          & 17 \\
Type=str & Dynamic (Type)                & 12 \\
\toprule
\end{tabular}
\caption{Feature predictive power as measured by ANOVA. F-score in
thousands to two significant figures; all p-values less than .01}
\label{tab-anova}
\end{table}

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
    ybar,
    symbolic x coords={noContext, noTypes, all},
    xtick=data,
    ymin=0,
    ylabel=Model Accuracy,
    enlarge x limits=0.5,
    legend style={at={(0.5,-0.15)},anchor=north}
]
    \addplot table[x=scoreName, y=score, col sep=comma]{removing-features-2.csv};
\end{axis}
\end{tikzpicture}
\caption{Accuracy when features are removed, based on decision trees on a
random subset of 20000 program pairs.}
\label{fig-removing-features}
\end{figure}

@@

\tabref{anova} gives an alternate view of the relative feature importances
as measured by ANOVA. In both tables we see that syntactic, dynamic, and
contextual features are all important.

We also try removing certain categories and retraining the model on what is
left. \figref{removing-features} shows that there is a significant decrease in
accuracy in each case, supporting the claim that all our feature categories are
important.

% prereq: define predictive power (leave-one-out, leave-one-in, Relief-F,
% ANOVA, whatever)

% Use as a rough guide: Table 3 on Page 8 of
% https://web.eecs.umich.edu/~weimerw/p/weimer-icsm2010.pdf

% Goal: Establish that we are smart for including all of these features. Note
% which individual features or feature categories were (not) included in
% previous work.

% Give a simple narrative about our effectiveness: implicitly, we were smart
% for deciding to include these features, better ingredients make better
% results, we have great results.

% TODO: what about turning on/off slicing?

% TODO: what about turning on/off types? (is that separate from slicing?)

@@

% PLOT LINES: line 1 is ``Nate programs'', line 2 is ``25\%, at random, of our dataset''

@@

In the most similar prior work, Learning to Blame, each program in the dataset
was a student's attempt at one of 23 different homework problems. Thus the
dataset was comparatively homogenous, making it unclear whether the results
would generalize to a more diverse dataset.

In our PythonTutor data by comparision, users wrote whatever programs they
wanted, so we
hypothesized that the data were more heterogeneous. To confirm this
quantitatively, we used agglomerative clustering both on our
data and on the Learning to Blame data; whichever dataset naturally splits
into more clusters is more heterogeneous.

% The exact number of clusters returned
% depends on the value we choose for the algorithm's threshold parameter, so we
% checked all thresholds at .01 intervals, starting at 0 and ending at the point
% where everything collapses into a single cluster. As shown in
% \figref{diversity}, regardless of the choice of threshold, our data breaks into
% more clusters than the NATE data. This confirms that our data is more
% heterogeneous.

To cluster programs, we transform the program ASTs into flat strings, and then
perform agglomerative clustering on them, with string edit distance as a distance
metric.
